Robust Bayes Reloaded

### 2023-01-19 papers/robayes/notes.tex
Add notes about robust Bayes

\title{Variana Reloaded}
\author{Alexis Roche}
%\date{}

@@

\section{Variational sampling recap}

Consider:
\begin{itemize}
\item a reference distribution~$\pi(x)$ (from which we can sample)
\item a factor~$f(x)$ so that $p\equiv \pi f$ is a distribution of interest
\item an approximating factor~$g$ searched under the exponential form:
  $$
  g(x) = e^{\theta^\top \phi(x)},
  $$ where $\phi$ is a known vector-valued function and $\theta$ needs
  to be determined.
\end{itemize}

VS is a stochastic method to approximate the parametric factor~$g$ that minimizes the generalized Kullback-Leibler divergence:
$$
D(\pi f\|\pi g)
= \int \pi (f \log \frac{f}{g} + g - f)
$$

It does so by sampling independent points $(x_1,\ldots,x_n)$ from~$\pi$ and minimizing a Monte Carlo approximation to the KL~divergence, leading to the surrogate criterion:
$$
\tilde{D}(\theta) = \frac{1}{n} \sum_i \ell[ f(x_i), g_\theta(x_i) ],
$$
where $\ell(f, g)$ is the loss function associated with the KL~divergence:
$$
\ell(f, g) \equiv f \log \frac{f}{g} - f + g
$$

We have proved in our initial paper from 2013 that the resulting estimator of the integral, or moment, $I(\phi) \equiv \int \pi f \phi$, is asymptotically unbiased with variance matrix:
$$
\Sigma_n = \frac{1}{n} \int \pi (f - g_\star)^2 \phi \phi^\top,
$$
where $g_\star$ is the genuine (intractable) KL~minimizer:
$$
g_\star \equiv \arg\min_g D(\pi f\|\pi g).
$$

This makes VS fundamentally different from importance sampling, for which the moment is directly estimated via:
$$
\hat{\mu} \equiv \frac{1}{n} \sum_i f(x_i)\phi(x_i),
$$
yielding an unbiased estimator with asymptotic variance:
$$
\Sigma_n^{IS} = \frac{1}{n} \left[
\int \pi f^2 \phi \phi^\top
- I(\phi) I(\phi)^\top
\right]
$$

In particular, VS has the key consistency property unlike IS to be exact when the target factor lies in the approximating family.

\section{Variance upper bound}

Intuitively, the closer the target factor~$f$ to the approximating family, the smaller the VS~asymptotic variance. To show that formally, we would like to upper bound the mean square error $\int\pi(f-g)^2$ by some increasing function of the KL~divergence $D(\pi f\|\pi g)$.

I have been bumping into this issue for years... only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit. We will need a straightforward intermediate result:

\paragraph{Alexis theorem:} if $f:\Omega\in\mathbb{R}\to \mathbb{R}$ is a smooth convex function from some open interval reaching its minimum in $a\in\Omega$, assuming for simplicity that $f(a)=0$, and such that the second derivative $f''$ is a decreasing function, then we have that:
$$
\forall x \leq a,
\qquad
f(x) \geq \frac{1}{2} f''(a) (x - a)^2 .
$$

The proof is straightforward from the fact that $f''(x) \geq f''(a)$ since $f''$ is decreasing, then we just need to integrate this relationship twice and take into account that $f'(a)=f(a)=0$.

Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
$$
\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
$$
implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
$$
\forall g\leq f,
\qquad
\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
$$

But we can also consider the loss as a function of~$f$ at fixed~$g$. We have:
$$
\frac{\partial\ell}{\partial f} = \log(f),
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{1}{f},
$$
which means that Alexis's theorem can be applied again, this time with $a=g$, yielding:
$$
\forall f\leq g,
\qquad
\ell(f, g) \geq \frac{1}{2g} (f - g)^2. 
$$

Concatenating both results, we see that for all $(f, g)$:
$$
(f-g)^2 \leq 2 \max(f, g)\ell(f, g).
$$

Integrating this over~$\pi$, we get this long-waited inequality:
$$
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
$$

Note that this is only tight if $\max(f,g)$ is constant,

This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes. But this holds under the mild condition that both $f$ and $g$ remain bounded. This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.

Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.

The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
$$
(f-g)\phi^2
\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
= 2 \max(f\phi, g\phi) \phi\ell(f, g)
= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
$$
where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
$$
\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
$$

This yields the same kind of variance upper bound for any moment estimator we have.

\section{VS vs IS}

The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$. This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor. The IS estimator is just a sample mean:
$$
\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
\qquad x_i \stackrel{iid}{\sim} \pi(x),
$$
which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
$$
\mathbb{E}[f(X)] = \int\pi f,
$$
we also have that:
$$
{\rm Var}[f(X)]
= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
$$

Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.

As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$. Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.

Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$? Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''. We can make this a little bit more precise using :
$$
\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
$$

\section{Custom optimizer}

We shall note that the basis functions $\phi_i$ underlying the fitting family are somewhat arbitrary, hence comes a natural question: which basis is most convenient for optimization?

Consider for now the case of the KL divergence $D(\theta)=D(\pi f\|\pi g_\theta)$ for a given set of $\phi$-functions. We have the relations:
$$
\nabla D = \int \pi (g_\theta-f) \phi,
\qquad
\nabla^2 D = \int \pi g_\theta \phi\phi^\top.
$$

From there, a key remark is that the Hessian at a particular $\theta$ is the identity matrix if $\phi$ defines an orthonormal basis for the dot product:
$$
\langle h_1, h_2 \rangle \equiv \int \pi g_{\theta} h_1 h_2,
$$
where $h_1$ and $h_2$ are real-valued functions... We have some Hilbert space structure here, but I'll skip these details for now as I am not a mathematician. What really matters is that the Hessian being the identity means that a Newton iteration is very easy to compute as it boils down to a first-order gradient iteration. Importantly, however, this property is achieved for a particular basis function that is {\em dependent on the current fit estimate}.

So, let's say that we have a current fit $g_n$ (which may be defined, for instance, by its coordinates in the canonical, fixed basis) and we want to perform one Newton iteration to improve it. Given an orthonormal basis $\phi_n$ associated with $g_n$ (in the sense of the above-defined dot product), we may look for the update under the form $g_\nu = g_n \exp({\nu^\top \phi_n})$. The Newton iteration from $\nu=0$ then takes the form:
$$
\hat{\nu} = -\lambda \nabla D(0) = \lambda \int \pi(f - g_n)\phi_n,
$$
for some damping factor $\lambda > 0$ (the genuine Newton iteration corresponds to $\lambda=1$), yielding the update:
$$
g_{n+1} = g_n \exp (\hat{\nu}^\top \phi_n).
$$

We will further need to represent $g_{n+1}$ in the same fixed coordinate system as $g_n$, i.e. using a ``$\theta$ parameter'', which will require a linear coordinate change from $\phi_n$ to the fixed basis.

\subsection{Orthonormal basis}

Let's start in one dimension for simplicity. Given a current fit $g$, let $q=\pi g$ (pig\'e?),  with integral, mean and scale parameters $(Z, m, s)$, the following three functions define an orthonormal basis:
$$
\phi_0(x) = \frac{1}{\sqrt{Z}},
\quad \phi_1(x) = \frac{(x-m)}{\sqrt{Z}s},
\quad \phi_2(x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x-m)^2}{s^2} - 1\right].
$$

{\bf Proof.} It is easy to see that $\|\phi_0\|=1$, $\|\phi_1\|=1$, $\langle \phi_0, \phi_1 \rangle=0$ and $\langle \phi_0, \phi_2 \rangle=0$. The reason why $\langle \phi_1, \phi_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric. Finally, note that:
$$
\phi_2(x)^2 = \frac{1}{2Z} 
\left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
$$
hence, given that the 4th-order centered moment of a Gaussian is~3 (which can be seen using the $\chi_1^2$ distribution), we have:
$$
\int q \phi_2^2 = \|\phi_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
$$

In multiple dimensions, we do:
$$
\phi_0(\x) = \frac{1}{\sqrt{Z}}, 
\quad
\phi_{1i}(\x) = \frac{(x_i-m_i)}{\sqrt{Z} s_i},
\quad
\phi_{2i}(\x) = \frac{1}{\sqrt{2Z}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
$$
This defines an orthonormal basis because the components of~$\x$ are mutually independent under the factorial distribution~$q(\x)$, hence for different basis vectors: 
$$
\forall (i,j), \quad i\not= j, \quad
\langle \phi_{ki}, \phi_{\ell j} \rangle
= \left(\int q \phi_{ki}\right)\left(\int q \phi_{\ell j}\right)
= \langle \phi_{ki}, 1 \rangle \langle \phi_{\ell j}, 1 \rangle = 0.
$$
Moreover, when it comes to dot products involving the same component, the one-dimensional results apply.

\subsection{Coordinate change}

Given some ``moving'' coordinates in the orthonormal basis, we aim to compute the corresponding coordinates in the canonical basis $b_0=1$, $b_{11}=x_1$, $b_{12}=x_1^2$, $b_{21}=x_2$, $b_{22}x_2^2$, etc. This is a linear mapping. We easily get by identification, owing to the separability of Gaussian factors:
$$
\phi_0 =
\frac{1}{\sqrt{Z}} b_0,
\qquad
\phi_{1i} =
-\frac{m_i}{\sqrt{Z}s_i} b_0
+ \frac{1}{\sqrt{Z}s_i} b_{1i},
\qquad
\phi_{2i} =
\frac{1}{\sqrt{2Z}} \left(\frac{m_i^2}{s_i^2} - 1\right) b_0
- \frac{\sqrt{2}m_i}{\sqrt{Z}s_i^2} b_{1i}
+ \frac{1}{\sqrt{2Z}s_i^2} b_{2i}.
$$

This shows that the transition matrix from moving to fixed coordinates is block-diagonal, up to a first row corresponding to the constant terms. In the 1D case, it is upper triangular. In general, it never has more than three nonzero elements per column.

$$
T =
\frac{1}{\sqrt{Z}}
\left[
\begin{array}{ccccccc}
1 & -\frac{m_1}{s_1} & -\frac{m_2}{s_2} & \ldots &
  \frac{1}{\sqrt{2}} \left(\frac{m_1^2}{s_1^2} - 1\right) &
  \frac{1}{\sqrt{2}} \left(\frac{m_2^2}{s_2^2} - 1\right) &
  \ldots \\
0 & \frac{1}{s_1} & 0 & \ldots &
-\frac{\sqrt{2}m_1}{s_1^2} & 0 & \ldots \\
  0 & 0 & \frac{1}{s_2} & 0 &
  0 & -\frac{\sqrt{2}m_2}{s_2^2} & \ldots \\
\vdots & \vdots & \vdots & \vdots &
\vdots & \vdots & \vdots\\
0 & 0 & 0 & 0 &
  \frac{1}{\sqrt{2}s_1^2} & 0 & \ldots\\
0 & 0 & 0 & 0 &
  0 & \frac{1}{\sqrt{2}s_2^2} & \ldots\\
\end{array}
\right].
$$

Let $\theta=T\nu$.

$$
\sqrt{Z}\theta_0 =
\nu_0
- \sum_i \frac{m_i}{s_i} \nu_{1i}
+ \frac{1}{\sqrt{2}} \sum_i \left(\frac{m_i^2}{s_i^2} - 1\right) \nu_{2i}
.
$$

$$
\sqrt{Z}\theta_{1i}
=
\frac{\nu_{1i}}{s_i} - \sqrt{2} \frac{m_i\nu_{2i}}{s_i^2}.
$$

$$
\sqrt{Z}\theta_{2i}
=
\frac{\nu_{2i}}{\sqrt{2}s_i^2}.
$$

\subsection{Application to VS}

In VS, we approximate the KL divergence using some integration rule with weights~$w_i$:
$$
\hat{D}(\nu)
\equiv \sum_i w_i \left[-f_i
  \log g_\nu(\x_i) + g_\nu(\x_i) \right],
$$
yielding the gradient (with simplified notation):
$$
\nabla\hat{D} =
\sum_i w_i (g_i - f_i) \phi(\x_i),
$$
where $\phi$ is the above defined orthonormal basis, which depends on the current fit~$g$.

Hence, the VS approximation to the Newton iteration on the KL~divergence is:
$$
\hat{\nu} = -\lambda \sum_i w_i (g_i - f_i) \phi(\x_i).
$$

This requires evaluating both the current fit and the associated orthonormal basis functions at each sample point. Evaluating the basis function requires computing the natural parameters $(Z, m, s)$. Note that, in the canonical basis, we don't have to re-evaluate the basis functions on each iteration, but we have to re-evaluate the KL Hessian if we are to implement the genuine Newton method.

Once the moving coordinates are updated through $\hat{\nu}$, we convert them to canonical coordinates by left multiplying $\hat{\nu}$ by the above transition matrix, which gives us the increment $\delta\theta=T\hat{\nu}$ to add to the current $\theta$ estimate for the next iteration.

\subsection{Utter bulshit}

So far in this section, we have considered the problem of minimizing the KL~divergence. We have seen that we can elegantly work around the problem of inverting the Hessian in a Newton iteration by using appropriate orthonormal bases. However, as we know, the KL gradient may be intractable for the target factor under consideration, meaning that we need to approximate it via random sampling or quadrature, leading respectively to stochastic gradient and numerical approximation methods.

There are essentially two options.

One option is to approximate the moment involved in the KL gradient:
$$
\nabla D \approx \int \pi g_n\phi - {\cal I}(\pi f \phi),
$$
where~${\cal I}$ denotes a numerical integration operator.

The other option is to approximate the KL gradient altogether:
$$
\nabla D \approx {\cal I}[\pi(g_n - f)\phi].
$$

Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family. Hence, for a near Gaussian target, we expect less numerical instabilities with Option~2. Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.

What we have sketched here is nothing but a stochastic gradient method to minimize the KL~divergence, which is quite different in spirit from the initial VS~idea in that resampling could take place at each iteration of the optimization process. While this ensures that the successive gradient estimates are mutually independent, hence the method is unbiased and exact (not only asymptotically). This could be costly if the target factor takes time to evaluate, though.

\section{Probability updating}

What the variance inequalities essentially prescribe is that the target factor~$f$ be a bounded function. If this condition holds, the product distribution $\pi f$ does not have heavier tails than $\pi$, therefore IS converges whenever $\pi \phi\phi^\top$ exists, which is guaranteed for instance if $\pi$ is Gaussian (proper) and  $\phi$ is polynomial, which is our standard setting. VS convergence follows from the same argument provided that $(f-g)$ is bounded, in other words that the best fitting~$g$ is also bounded.

polynomial in the Gaussian case. In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function. Other compliant factors can be derived from this case, as we will see. However, importance weighting functions resulting from distribution ratios are ruled out.

Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function. It is natural to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated. This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior. Makes sense. This is the canonical approach exposed in my previous publications on VS.

However, there are some issues with that. Firstly, the likelihood might not be fast-decreasing (this is the case in image registration), meaning that VS convergence is doubtful. It may not converge at all, or convergence may be super slow... But, even if $L$ is a nice friendly sharply peaked function, sampling from the prior is intuitively not the most efficient thing to do as most simulated points will have low likelihood.

This is what we may call the ``grasp all, loose all'' effect: unless the factor is stricly Gaussian, which never happens in practice (and if it did, VS would not be useful), the VS~variance blows up as the context gets wider. In a narrow context, i.e. $\pi$ is sharp wrt $f$, the factor doesn't vary much within the sampled region, hence it can be very well approximated by a Gaussian (think of the fact that $\log f$ may be close to a quadratic function in this small neighborhood by Taylor's expansion). Such an approximation is bound to be less accurate as the context gets wider.

\section{Variance reduction}

So, how can we make VS more stable?

Instead of trying to solve the problem in one shot, maybe we can do it iteratively. Assume we have a current factor approximation $g_n$. How can we improve it?

Perhaps a natural idea, at least this is among the first that came to my mind when I started to work on VS, is to apply VS sampling from~$\pi g_n$ to the factor $f/g_n$. In my initial framework, it means approximating the target $p=\pi f$ by the distribution $q=\pi g$, using the current approximation $q_n$ as a sampling distribution and $p/q_n$ as the target importance weight function.

But we know that this is a bad idea because $p/q_n=f/g_n$ may not be upper bounded, hence VS is not even guaranteed to converge. In fact, it will diverge if $p$ is a heavy-tailed distribution, which is impossible to know in advance.

A more reasonable thing to do is to note that, since $g_n$ approximates $f$, the combined function $g_n^{1-\alpha}f^\alpha$, for some $0< \alpha\leq 1$, yields an even better approximation. If $g_n$ is far off, it would be a little less far off if we pick a small~$\alpha$, and much closer if we pick a value close to one. If $g_n$ is already a good fit, it shouldn't matter too much which $\alpha$ we pick, at least it should have less impact.

This simple remark suggests that the objective $D(\pi f\|\pi g)$ could be proxied by another KL divergence, $D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)$, yielding the following iteration:
$$
g_{n+1} = \arg\min_g D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)
$$

We will soon realize that this is heavily related to the minimization of the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$, and also has a strong connection with the Expectation Propagation algorithm, more specifically its ``averaged'' variant (AEP).

But, for the time being, let us see why this is interesting for VS.

\section{Scaling}

We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$. This second interpretation opens a new perspective on~VS, and enables generalizations. For instance, we may consider the more general divergence:
$$
\delta_\varepsilon (f, g) \equiv
D(\pi f^\varepsilon \|\pi g^\varepsilon),
$$
where $\varepsilon$ is a positive scaling factor. The intuition behind this generalization is that the scaled target factor $f^\varepsilon$ gets flatter as $\varepsilon$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions. On the other hand, applying a scaling factor~$\varepsilon$ smaller than one amounts to ``shrinking'' the context with respect to the factor, hence biasing the divergence minimizer (the VS target) towards the particular context. This yields a typical bias/variance tradeoff.

Could the Expectation Propagation (EP) solve this?

$$
p = \pi \prod_{k=1}^n f^\varepsilon,
\quad
\varepsilon = \frac{1}{k}.
$$

Say current approx is $g^\epsilon$ for all factors. An EP update is to fit the factor $f^\varepsilon$ in the cavity $c=\pi g^{1-\varepsilon}$ by minimizing $D(cf^\varepsilon\|cg^\varepsilon)$ over a new $g$. This can be approximated using VS, with $c$ as the context, and $f^\varepsilon$ as the target factor. This is arguably an easier problem than the original one.

\section{Expectation Propagation}

We propose to see online learning as an iterative fitting procedure in which each iteration performs an update of the form:
$$
q \leftarrow F(q^{1-\alpha} p^\alpha),
$$
where $F$ is a fitting operator, $q$ is the current objective approximation, $p$ is an empirical target distribution (possibly based on a random mini-batch), and $0\leq \alpha\leq 1$ is a learning rate.

%If $\alpha=1$, this is in fact classical optimization applied to~$h$ (assuming the same one across iterations). This shows that offline learning is just a special case of online learning.

This general iteration may be seen as a stochastic version of message passing for learning, as proposed in my amazing 2016 paper \cite{rr:16}, which treats mini-batches in a deterministic manner and enables factor-wise approximations, but requires heavy memory load. Conceptually, it's pretty much the same story but the stochastic version is much more feasible. In the message passing framework, the target is of the form $p=cf$, with $c=q^{1-\alpha}$ defining the ``context'' and $f=p^\alpha$ defining the ``factor'' to be approximated, hence the whole story is about fitting factors and the learning specifics are brushed under the carpet.

The iteration is doubly approximative: the fitting objective at each iteration is an approximation, and the fitting procedure itself is approximative. How damaging is it? If the objective is exact, then the only thing we need is a good fitting scheme. If the fitting works well, then the only constraint is the limited batch size, which forces us to approximate the objective iteratively.

We shall consider these as {\em separate problems}. Would it hurt to have an exact fitting scheme? It would not, for sure. We will focus here on minimizing (generalized) KL divergence as a fitting strategy:
$$
F_\star(p, q) = \arg\min_q D(p\|q),
$$
hence defining our {\em ideal} fitting operator. If the target is of the form $p=wf$, this may be interpreted as minimizing the local KL divergence $D(wf\|wg)$ along~$g$, so it can be seen as factor approximation in the sense of the local KL divergence. But we may also consider that we are minimizing an adaptive target {\em globally}.

Assume for now that we are dealing with a fixed target~$p$. Ideally, we would set $\alpha=1$ to fit~$p$ as quickly as possible. However, the fit can only be approximated in practice and may be poor if the target is far off the approximating family. A target of the form $q^{1-\alpha}p^\alpha$ is intuitively `easier' to fit if~$q$ lies in the approximating family.

%On the other hand, for a narrow context, VS will pretty much behave like a Taylor expansion. So, if we have to choose between the two options, we could say we are on the safe side using a narrow context, but the reality is that we may get a poor fit in both cases.

%We may assume that the best option is a context that is more or less centered at the target peak and just wide enough to capture relevant variations. If the target is close to the approximation space, then it does not matter much whether~$w$ is narrow or wide. But if it's not, then the question is which scale is relevant to describe it. The global KL fit would certainly help. Problem is: it's exactly what we are looking for. Chicken and egg problem.

The proposed scheme could be interpreted as gradual `non-gaussianity' or some sort of bridge sampling. Say we start with a very small~$\alpha$ and some prior guess~$q=\pi$. Since~$\alpha$ is small, our target $\pi^{1-\alpha}p^\alpha$ at first iteration is almost Gaussian. We get an update~$q$ that is just a little skewed towards~$p$:
$$
q \approx \pi^{1-\alpha} p^\alpha
$$

Now, if we update the target according to $p=q^{1-\alpha}p^\alpha$ and re-fit it, the new fit is a little bit more skewed towards~$p$. Roughly,
$$
q^{1-\alpha} p^\alpha 
\approx \pi^{(1-\alpha)^2} p^{(1-\alpha)\alpha + \alpha}
= \pi^{1-\beta} p^{\beta},
$$
with $\beta=\alpha+(1-\alpha)\alpha$ evaluating the larger weight of~$p$ relatively to the prior~$\pi$ in the new target, $\beta=\alpha+(1-\alpha)\alpha\approx 2\alpha$.

%The updated~$q$ is a natural choice for the context since it is, in principle, a closer fit to~$p$ than the prior.

So the idea is pretty simple: pick a fixed value~$\alpha>0$ and iterate the process of fitting $q^{1-\alpha}p^\alpha$. After several (many?) iterations, we hope that the target closely resembles~$p$. Is that naive?

What can go wrong? Well, if we use too large a value for~$\alpha$, this clearly won't work... So one first question is: how do we make sure that~$\alpha$ is small enough?

\subsection{Consistency}

Let~$p$ a target distribution and $Q$ an approximating family. A sufficient condition for the KL fit (in an exponential family of the type considered in the seminal paper) to have finite integral is obviously that~$p$ have a finite integral because $\int q_\star = \int p$. Moreover, if~$p$ has a finite integral, the same goes for $wp$ for any bounded function~$w$, therefore the contextual KL fit will always satisfy $\int q < \infty$. This implies that, if the target distribution is proper, the iterative KL minimization considered here yields a proper distribution.

\subsection{Fixed point}

Assume that~$Q$ is an exponential family $\exp(\theta^\top \phi)$. The fixed point~$q_*$ of our algorithm is such that:
$$
q_* = \arg\min_q D(q_*^{1-\alpha} p^\alpha \| q),
$$
therefore it satisfies the moment condition
$$
\int q_*^{1-\alpha}p^\alpha\phi 
= \int q_* \phi
$$

Consider the $\alpha$-divergence:
$$
D_\alpha(p\|q) = 
\frac{1}{\alpha(1-\alpha)}
\int \alpha p + (1-\alpha) q - p^\alpha q^{1-\alpha} 
$$

Consider minimizing it on~$Q$. Whenever $\alpha\not=0,1$, the first order optimality condition reads:
\begin{eqnarray*}
\alpha(1-\alpha) \frac{\partial D_\alpha}{\partial \theta} 
& = & \int (1-\alpha) q \phi - p^\alpha (1-\alpha) q^{1-\alpha} \phi \\
& = & (1-\alpha) \int (q - p^\alpha q^{1-\alpha}) \phi \\
& = & 0
\end{eqnarray*}

Incredible!!! I am realizing now that Minka \cite{Minka-05} noted this result before, but maybe not the implication: it means that the $\alpha$-divergence optimality condition is exactly the same as for the iterative scheme considered here using a {\em fixed target}, call it simple EP or whatever you like.  Since $D_\alpha$ is convex, this means that, whenever simple EP converges, it converges to the unique minimizer of $D_\alpha(p\|q)$ on~$Q$.  But how comes Minka or \cite{Dehaene-18} did not include this result in their convergence analysis of EP or AEP? Because they were using EP or AEP with variable targets in a data partitioning perspective.

The remaining issue is convergence. Basically, AEP boils down to iterated $M$-projections:
\begin{equation}
\label{eq:iter_m_proj}
q \leftarrow \arg\min_{q'} D(q^{1-\alpha}p^\alpha\|q').
\end{equation}

Iterated projection is a typical algorithm to find the intersection point between two sets, but AEP is not quite the same thing. There is only one projection set here and, at each iteration, the current projection is slightly pushed away before re-projecting.

We know that, if convergence occurs, then AEP minimizes $D_\alpha(p\|q)$, but is AEP guaranteed to converge? It would help to have a monotonicity property, {\em i.e.}, the update~(\ref{eq:iter_m_proj}) always decreases $D_\alpha(p\|q)$. Is it true? Minka \cite{Minka-05} does not seem to think so.

In practice, anyway we can only approximate the $M$-projection, so there might be no theoretical convergence guarantee. It may thus be wise to try and modify the iterative scheme so as to force convergence. A similar situation occurs in gradient descent schemes, which are not guaranteed to converge unless the step size parameter is tuned on each iteration so as to make the objective decrease.

Can we force convergence? Well, we have a fixed point iteration:
$$
\theta_{n+1} = f(\theta_n),
$$
with:
$$
f(\theta) = \arg\min_{\theta'} \tilde{D}(q_\theta^{1-\alpha}p^\alpha\|q_{\theta'}),
$$
where $\tilde{D}$ is some tractable KL divergence proxy. The function $f(\theta)$ is fully determined by~$p$ and~$\alpha$. The update may be generalized using a damping factor or ``learning rate'' $0<\epsilon_n\leq 1$:
$$
\theta_{n+1} = (1-\epsilon_n)\theta_n + \epsilon_n f(\theta_n).
$$

% We can always tune $\epsilon_n$ so that $\|\theta_{n+1}-\theta_n\|\leq \|\theta_n-\theta_{n-1}\|$, thereby ensuring that $\theta_n$ is a Cauchy sequence.

In practice, a small constant value for~$\epsilon$ might do. In a safer version of the algorithm, we may tune $\epsilon_n$ on each iteration so as to guarantee that some $\alpha$-divergence proxy decreases.

\subsection{AEP as relaxed KL~minimization}

The fact that fixed-target AEP is a convex minimization algorithm is certainly an interesting result that qualifies it as a relaxation method, which replaces the KL divergence objective with another $\alpha$-divergence for $0<\alpha<1$, corresponding to an {\em exclusive} and {\em non-zero forcing} divergence.

Why performing such a relaxation? Because the original problem is intractable. That is the only valid reason. By adopting this relaxation, we assume that we are able to minimize the KL divergence at least approximately when the target is `close enough' to the approximating family. So the existence of an approximate KL minimization scheme is at the heart of this strategy.

What are the candidates? 
\begin{enumerate}
\item Laplace method.
\item Approximate integration (remarking the equivalence between KL min and moment matching), so anything from the quadrature and Monte Carlo literature.
\item Variational sampling. (We could include in this category approaches such as Bayesian Monte Carlo, i.e. VS variants that make use of other costs than the empirical KL divergence, but such methods do not make much sense in the context of approximate KL minimization, do they?). 
\end{enumerate}

We could dismiss approximate integration from the very beginning based on a consistency argument: the fact that they are inexact even if the target lies in the approximating family, unlike both Laplace and VS. So, we have two remaining candidates.

What are the compared pros and cons of Laplace and VS? I can think about the following:
\begin{itemize}
\item[$-$] Laplace is potentially more efficient if the gradient and hessian have closed-form expressions.
\item[$+$] Laplace is local: it won't work well if the target is `rugged' or skewed, while VS is arguably more robust to such features. 
\item[$+$] Laplace does not handle singularities unlike VS (think about a... Laplace distribution, for which the Hessian does not exist at the mode).
\item[$+$] Laplace can output non-positive hessians, while VS can be constrained not to.
\end{itemize}

To put it in Christopher Bishop's terms: ``The most serious limitation of the Laplace framework, however, is that it is based purely on the aspects of the true distribution at a specific value of the variable, and so can fail to capture important global properties.'' In short, Laplace is potentially faster but less accurate than VS. At least, this is what I could have shown experimentally in my introducing paper, but I decided instead to consider VS as a moment matching method... how clever.

%We can't do it when it is far off, but can do it if is close enough -- how weird does that sound? I don't know. Arguably, it is easier to minimize the $\alpha$-divergence for small~$\alpha$ than for $\alpha\approx 1$. I am not even talking about $\alpha\leq 0$ or $\alpha > 1$, fuck it.

\section{Bridge sampling scheme}

Consider the series:
$$
\alpha_{n+1}=\alpha_n+(1-\alpha_n)\alpha_n=(2-\alpha_n)\alpha_n,
\qquad
0 < \alpha_0 < 1 .
$$

Clearly, $\alpha_n$ is increasing, upper bounded and therefore converges to a limit satisfying $\ell=(2-\ell)\ell$, hence $\ell=1$ cause $\ell=0$ is impossible if $\alpha_0>0$.

Let $\beta=(2-\alpha)\alpha$. We note that: $1-\beta=1-\alpha - \alpha(1-\alpha)=(1-\alpha)^2$.

\section{Laplace approximation}

Conjecture: if~$w$ approaches a Dirac distribution, the KL fit is the Laplace approximation, {\em i.e.} the fit derived from the Taylor expansion at the cavity center. Is that true? Sounds obvious. What else could it be?

If it's true, then we can claim that gradient descent and Newton method are just limiting cases of local KL minimization. If it's not, then KL minimization might raise some suspicion.

\section{Gradient descent}

One simple and classical way to perform fitting on each iteration is via a second-order Taylor expansion of the target logarithm:
$$
F = \log t = (1-\alpha)\log q + \alpha \log p.
$$

In the `quick Laplace' approach \cite{rr:16}, we perform the expansion at the cavity center~$a$, which verifies $\nabla \log q(a)=0$, hence:
$$
\nabla F(a) = \alpha g_p(a),
\qquad
\nabla\nabla^\top F(a) = (1-\alpha)H_q + \alpha H_p(a)
$$

Assuming that~$H$ is positive definite, the value that minimizes the local quadratic approximation is $b = a - H^{-1}g$, hence the update formulas:
$$
H \leftarrow (1-\alpha)H + \alpha H_p(a),
\qquad
a \leftarrow a - \alpha H ^{-1}g_p(a).
$$

>>>>

\title{Robust Bayes Reloaded}
\author{Alexis Roche}
%\date{}

@@

\section{Initial toughts}

The Bayesian paradigm is semi-generative in essence in that it relies on a parametric distribution model that incorporates a priori on unknown parameters. This a priori is non-testable in essence, and can thus be chosen in an arbitrary manner, which may lead to really poor inference models for high-dimensional models... This is an undeniable {\em weakness} of classical Bayesian analysis.

Robust Bayes analysis can be used to work around this problem. We may see it as a minimax risk strategy (instead of Bayesian) to select a reasonably low-dimensional parametric model to fight the curse of dimensionality. It is, in essence, a dimension reduction scheme.

Note that we are not talking about overfitting, at least in a traditional sense.
